# Submission for "Reproducible Research: Peer Assessment 1"

## Introduction

With the proliferation of personal activity monioring devices, it is now possible to get data on personal movements and behaviour. This exercise uses R to explore the data collected from a personal activity monitoring device which collects data at 5 minute intervals through out the day. The dataset consists of two months of data collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading and preprocessing the data
First of all, the data is read using read.csv function.

```{r eval=TRUE}
act <- read.csv("data/activity.csv", header=TRUE, na.strings= "NA",
                  colClasses=c("integer", "Date", "integer"))
```

The dataset has `r nrow(act)` rows and `r length(act)` columns with the following column names: `r names(act)`.

## Exploring mean total number of steps taken per day
The following code uses plyr to summarize the data by calculating sum of number of steps for each day.

```{r eval=TRUE}
library(plyr)
stepsPerDay <- ddply(act, c("date"), function(xdf) {
  sum <- sum(xdf$steps, na.rm=TRUE)
  return(data.frame(sum))})
```

This gives the dataframe ```stepsPerDay``` with sum of total steps taken for each day in the dataset.

```{r eval=TRUE}
head(stepsPerDay,8)
```

The following code calculates the mean and median total number of average steps per day.

```{r eval=TRUE}
meanSteps <- mean(stepsPerDay$sum)
medianSteps <- median(stepsPerDay$sum)
print(list(mean = meanSteps, median = medianSteps))
```

The following code plots the histogram of the total number of steps taken each day using ```ggplot2``` function.

```{r eval=TRUE, fig.height=6, fig.width=10}
library(ggplot2)
hplot1 <- ggplot(stepsPerDay, aes(x=sum)) +
  geom_histogram(binwidth=1000, colour="black", fill="lightgrey")
hplot1 <- hplot1 + geom_vline(data=stepsPerDay, aes(xintercept=mean(sum, na.rm=T)),
                             size=1, linetype=2, colour="blue") +
  geom_vline(data=stepsPerDay, aes(xintercept=median(sum, na.rm=T)), size=1, linetype=2, colour="red") +
  geom_text(aes(mean(sum, na.rm=T)-400, 7.5, label=paste0("Mean = ",
                                                          as.character(round(mean(sum, na.rm=T)),2))),
              angle=90, color="blue", size=4,
              data=stepsPerDay) +
    geom_text(aes(median(sum, na.rm=T)-400, 7.5, label=paste0("Median = ",
                                                          as.character(round(median(sum, na.rm=T)),2))),
              angle=90, color="red", size=4,
              data=stepsPerDay) +
  xlab("Total number of steps") + ylab("Count") + theme(legend.justification = 'right', legend.position=c(0.25,0.3)) +
  scale_fill_discrete(name="Experimental\nCondition") +
  ggtitle("Histogram of total steps per day") + theme_bw()
hplot1
```

## Exploring the average daily activity pattern

The following code plots a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days. First, the data is summarized by total number of steps for each interval using ```ddply``` function. 

```{r eval=TRUE, fig.height=6, fig.width=8}
stepsPerInterval <- ddply(act, c("interval"), function(xdf) {
  sum <- sum(xdf$steps, na.rm=TRUE)
  mean <- mean(xdf$steps, na.rm=TRUE)
  median <- median (xdf$steps, na.rm=TRUE)
  return(data.frame(cbind(sum, mean, median)))})
```

The stepsPerInterval dataframe has the summary of the number of steps, mean steps and median steps for each interval.

```{r eval=TRUE, fig.height=6, fig.width=8}
head(stepsPerInterval,8)
```

The following code plots the average number of steps for the intervals averaged across all the days available.


```{r eval=TRUE, fig.height=6, fig.width=8}
with(stepsPerInterval, plot(interval, mean, type="n", main="Average daily steps pattern", ylab="Steps",
                       xlab="Date"))
with(stepsPerInterval, lines(interval, mean))
with(stepsPerInterval, abline(v=stepsPerInterval[stepsPerInterval$sum == max(stepsPerInterval$sum), c(1)],
                              col="red"))
library(xts)
library(scales)
data.xts <- as.xts(1:288, as.POSIXct("2014-01-01 00:00", tz = "GMT") + 60 * 5 * (0:287))
stepsPerInterval$timeInterval <- index(data.xts)
p.intervals <- ggplot() + geom_line(data=stepsPerInterval,
                               aes(timeInterval, mean),
                               stat="identity", alpha=1, size=.7, col="brown") +
  ggtitle(paste("Daily activity pattern")) +
  scale_x_datetime(breaks = "1 hour", labels=date_format("%H:%M:%S")) +
  xlab("5 minute intervals") + ylab("average number of steps") +
  theme_bw() + theme(legend.position="none", axis.text.x = element_text(angle = 90, hjust = 1))
p.intervals
```


The following code calculates particular 5-minute interval which contains the maximum number of steps.

```{r eval=TRUE, fig.height=6, fig.width=8}
stepsPerInterval[stepsPerInterval$sum == max(stepsPerInterval$sum),]
```
As shown in the above output, the maximum number of average steps is taken at interval 835 i.e. from 8:35am to 8:40am interval.

## Imputing missing values

The following code calculate the total number of missing values in the dataset.

```{r eval=TRUE}
sum(is.na(act$steps))
```

The following code imputes the NA values with average number of steps across the dataset for that particular time interval. The new dataset ```act2``` contains the steps with NAs imputed using the avrage numbers for the particular time interval.

```{r eval=TRUE}
getMeanSteps <- function(xinterval) {
  value <- NULL
  for (i in 1:length(xinterval))
  {
    value <- c(value, stepsPerInterval[stepsPerInterval$interval == xinterval[i], c(3)])
  }
  return(value)
}
act2 <- act
nax <- which(is.na(act2$steps))
act2$steps[nax] <- getMeanSteps(act2$interval[nax])
tail(act2)
```

 
This gives the data with sum of steps per day.

```{r eval=TRUE}
stepsPerDay2 <- ddply(act2, c("date"), function(xdf) {
  sum <- sum(xdf$steps, na.rm=TRUE)
  return(data.frame(sum))})
```

and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r eval=TRUE}
head(stepsPerDay2,10)
```
Calculate mean and median
```{r eval=TRUE}
#mean
meanSteps <- mean(stepsPerDay2$sum)
#median
medianSteps <- median(stepsPerDay2$sum)
print(list(mean = meanSteps, median = medianSteps))
```

The following code creates a histogram of the total number of steps taken each day.

```{r eval=TRUE, fig.height=6, fig.width=8}
require(ggplot2)
hplot2 <- ggplot(stepsPerDay2, aes(x=sum)) +
  geom_histogram(binwidth=1000, colour="black", fill="lightblue") +
  #geom_density(alpha=.2, colour="red3") +
  geom_vline(data=stepsPerDay2, aes(xintercept=mean(sum, na.rm=T)), size=1, linetype=2, colour="blue") +
  geom_vline(data=stepsPerDay2, aes(xintercept=median(sum, na.rm=T)), size=1, linetype=3, colour="red") +
  xlab("Total number of steps") + ylab("Count") +
  ggtitle("Histogram of total steps per day")
hplot2
```

## Are there differences in activity patterns between weekdays and weekends?

To check differences in activity pattern during weekdays and weekends, the data needs to be classified into two categories - weekend or weekday. The following code adds a ```weekend``` column to the act2 dataframe which has a string weekend or weekday depending on wheather the date is a weekend or a weekday respectively.

```{r eval=TRUE, fig.height=6, fig.width=8}
##function to check if a date is a weekend or a weekday
## the function is not named as a verb in line with the is.{condition} format as in
## is.na or is.data.frame
is.weekend <- function(xdate) {
  if (weekdays(xdate) == "Saturday" | weekdays(xdate) == "Sunday") {return("Weekend")} else
  {return("Weekday")}
}
act2$weekend <- sapply(act2$date, is.weekend)
act2$weekend <- factor(act2$weekend)
```

The following code plots the activity data by weekend variable.

```{r eval=TRUE, fig.height=6, fig.width=8}
stepsPerIntervalWeek <- ddply(act2, c( "weekend","interval"), function(xdf) {
  sum <- sum(xdf$steps, na.rm=TRUE)
  mean <- mean(xdf$steps, na.rm=TRUE)
  median <- median (xdf$steps, na.rm=TRUE)
  return(data.frame(cbind(sum, mean, median)))})
stepsPerIntervalWeek$timeInterval <- index(data.xts)
plot3 <- ggplot() + geom_line(data=stepsPerIntervalWeek,
                               aes(timeInterval, mean),
                               stat="identity", alpha=1, size=.7, col="brown") +
  ggtitle(paste("Differences in daily activity trend - weekday vs. weekend")) +
  scale_x_datetime(breaks = "1 hour", labels=date_format("%H:%M:%S")) +
  xlab("5 minute intervals") + ylab("average number of steps") +
  theme_bw() + theme(legend.position="none", axis.text.x = element_text(angle = 90, hjust = 1)) + facet_wrap(~weekend, ncol=1)
plot3
```

From the above graph, we can compare the daily activity trends during weekdays and weekends. 



